`mathml`: Translate R Expressions to MathML and LaTeX

### 2023-01-30 rj/mathml.Rmd
acknowledgement
initial checks

---
title: "mathml: Translate R expressions to MathML and LaTeX"
date: "2023-01-28"
abstract: >
  This R\ package translates R\ objects to suitable elements in MathML or LaTeX,
  thereby allowing for a pretty mathematical representation of R\ objects and
  functions in data analyses, scientific reports and interactive web content. In
  the RMarkdown document rendering language, R\ code and mathematical content
  already exist side-by-side. The present package allows to use the same
  R\ objects for both data analysis and typesetting in documents or web content.
  This tightens the link between the statistical analysis and its verbal
  description or symbolic representation, which is another step towards
  reproducible science. User-defined hooks allow to extend the package by
  mapping specific variables or functions to new MathML and LaTeX entities. A
  number of examples and a case study illustrate the use of the package in a
  scientific report.
draft: true
author:  
  - name: Matthias Gondan
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: bibliography.bib
preamble:
- \usepackage{cancel}
- \usepackage{amsmath}
Vignette: >
  %\VignetteIndexEntry{mathml: Translate R expressions to MathML and LaTeX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

@@

The paper is organized as follows. Section\ 2 presents the technical background
of the package, including the two main classes of Prolog rules for translating
R\ objects to mathematical expressions. Section\ 3 illustrates the main features
of the \pkg{mathml} package, potential issues and workarounds using examples
from the day-to-day perspective of a user. Section\ 4 present a case study for
the use of the package in a scientific report. Section\ 5 concludes with a
discussion and ideas for further development.

@@

The translation of R\ expressions to mathematical output is achieved through a
Prolog interpreter from R\ package \pkg{rolog} [@rolog]. Prolog is a classical logic
programming language with many applications in expert systems, computer
linguistics and symbolic artificial intelligence. The main strength of Prolog is
its concise representation of facts and rules for knowledge and grammar, as well
as its efficient built-in search engine for closed world domains. As it is
well-known, R\ is a statistical programming language for data analysis and
statistical modeling. Whereas Prolog is weak in statistical computation, but
strong in symbolic manipulation, the converse may be said for the R\ language.
The \pkg{rolog} package bridges this gap by providing an interface to a SWI-Prolog
distribution [@swipl] in R. The communication between the two systems is mainly
in the form of queries from R\ to Prolog, but two predicates allow Prolog to
ring back and evaluate terms in R.

@@

term <- quote(NaN + NA + TRUE + FALSE + Inf + (-Inf))
mathout(term)
```

@@

To avoid name clashes with package \pkg{stats}, `roof()` is used to put a hat
on a symbol (see next section for further decorations). Note that an 
R\ function `roof()` does not exist in base R, it is provided by the package
for convenience and points to the identity function.

@@

For standard operators with known precedence, \pkg{mathml} is generally able to
detect if parentheses are needed; for example, parentheses are automatically
placed around `g + h` in the `nodot`-example. However, we note unecessary
parentheses around `l + m` above. Thes parentheses are a consequence
of `quote(a^(b + c))` actually producing a nested R\ call of the
form `'^'(a, (b + c))` instead of `'^'(a, b + c)`:

@@

A few more examples are shown in Table \@ref(tab:base-stats), including
functions from \pkg{stats}.

@@

Note that in the example above, the quoted term is an abbreviation
for `call("integrate", quote(sin), ...)`, with `sin` being an R\ symbol, not a
function. While the R\ function `integrate()` can handle both symbols and
functions, \pkg{mathml} needs the symbol because it is unable to determine the
function name of custom fuctions.

@@

Note that the seemingly more convient `term <- quote(A + B)` yields \(A + B\) in
the output---instead of the desired matrix representation. This behavior is
expected because quotation of R calls also quote the components of the
call (here, _A_ and _B_).

@@

Further customizations require the assertion of custom Prolog rules `math/2`,
`ml/3`, `jax/3`, as shown in the introduction.

@@

For asynchronous stimuli, Schwarz [-@schwarz1994] derived the expected
first-passage time as a function of the stimulus onset asyncrony \(\tau\),

@@

Assuming that the model is correct, the observable mean reaction times follow an
approximate Normal distribution around the model prediction \(E(T_\tau)\) for
each condition. We can, therefore, use a standard goodness-of-fit measure
by \(z\)-standardisation.

@@

In its current version \pkg{mathml} has some limitations. For example, it is
currently not possible to use the functions of \pkg{mathml} for writing
intratextual formulas, here, the user has to adopt the usual LateX notation.
Moreover, for long equations, we did not yet find a convenient way to insert
linebreaks. This is mostly due to lacking support by MathML and LaTeX renderers.
For example, in its current stage, the LaTeX package \pkg{breqn} [@breqn] is mostly
a proof of concept.

@@

We assume that `nthroot(x, n)` is available in the current
namespace [manually defined, or from R package \pkg{pracma}, @pracma], so that
the names of the arguments and their order are accessible to `canonical()` if
needed. As we can see below, \pkg{mathml} uses a default
representation `name(arguments)` for such unknown functions.

@@

A proper MathML term is obtained by `mlx/3` (the x in mlx indicates that it is
an extension and is prioritized over the default ml/3 rules). `mlx/3`
recursively invokes `ml/3` for translating the function arguments _X_ and _N_,
and then contructs the correct MathML entity `<mroot>...</mroot>`.

@@

>>>>

---
title: "`mathml`: Translate R Expressions to MathML and LaTeX"
date: "2023-01-28"
abstract: >
  This R\ package translates R\ objects to suitable elements in MathML or LaTeX,
  thereby allowing for a pretty mathematical representation of R\ objects and
  functions in data analyses, scientific reports and interactive web content. In
  the RMarkdown document rendering language, R\ code and mathematical content
  already exist side-by-side. The present package allows to use the same
  R\ objects for both data analysis and typesetting in documents or web content.
  This tightens the link between the statistical analysis and its verbal
  description or symbolic representation, which is another step towards
  reproducible science. User-defined hooks allow to extend the package by
  mapping specific variables or functions to new MathML and LaTeX entities. A
  number of examples and a case study illustrate the use of the package in a
  scientific report.
draft: true
author:  
  - name: Matthias Gondan
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: bibliography.bib
preamble:
- \usepackage{cancel}
- \usepackage{amsmath}
Vignette: >
  %\VignetteIndexEntry{mathml: Translate R expressions to MathML and LaTeX}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

@@

The paper is organized as follows. Section\ 2 presents the technical background
of the package, including the two main classes of Prolog rules for translating
R\ objects to mathematical expressions. Section\ 3 illustrates the main features
of the \CRANpkg{mathml} package, potential issues and workarounds using examples
from the day-to-day perspective of a user. Section\ 4 present a case study for
the use of the package in a scientific report. Section\ 5 concludes with a
discussion and ideas for further development.

@@

The translation of R\ expressions to mathematical output is achieved through a
Prolog interpreter from R\ package \CRANpkg{rolog} [@rolog]. Prolog is a classical logic
programming language with many applications in expert systems, computer
linguistics and symbolic artificial intelligence. The main strength of Prolog is
its concise representation of facts and rules for knowledge and grammar, as well
as its efficient built-in search engine for closed world domains. As it is
well-known, R\ is a statistical programming language for data analysis and
statistical modeling. Whereas Prolog is weak in statistical computation, but
strong in symbolic manipulation, the converse may be said for the R\ language.
The \pkg{rolog} package bridges this gap by providing an interface to a SWI-Prolog
distribution [@swipl] in R. The communication between the two systems is mainly
in the form of queries from R\ to Prolog, but two predicates allow Prolog to
ring back and evaluate terms in R.

@@

term <- quote(round(3.1415, 3L) + NaN + NA + TRUE + FALSE + Inf + (-Inf))
mathout(term)
```

@@

To avoid name clashes with package `stats`, `roof()` is used to put a hat
on a symbol (see next section for further decorations). Note that an 
R\ function `roof()` does not exist in base R, it is provided by the package
for convenience and points to the identity function.

@@

For standard operators with known precedence, \pkg{mathml} is generally able to
detect if parentheses are needed; for example, parentheses are automatically
placed around `g + h` in the `nodot`-example. However, we note unnecessary
parentheses around `l + m` above. Thes parentheses are a consequence
of `quote(a^(b + c))` actually producing a nested R\ call of the
form `'^'(a, (b + c))` instead of `'^'(a, b + c)`:

@@

A few more examples are shown in Table \@ref(tab:base-stats), including
functions from `base` and `stats`.

@@

Note that in the example above, the quoted term is an abbreviation
for `call("integrate", quote(sin), ...)`, with `sin` being an R\ symbol, not a
function. While the R\ function `integrate()` can handle both symbols and
functions, \pkg{mathml} needs the symbol because it is unable to determine the
function name of custom functions.

@@

Note that the seemingly more convenient `term <- quote(A + B)` yields \(A + B\)
in the output---instead of the desired matrix representation. This behavior is
expected because quotation of R calls also quote the components of the
call (here, _A_ and _B_).

@@

Further customization requires the assertion of new Prolog rules `math/2`,
`ml/3`, `jax/3`, as shown in the Appendix.

@@

For asynchronous stimuli, Schwarz [-@schwarz1994] derived the expected
first-passage time as a function of the stimulus onset asynchrony \(\tau\),

@@

Assuming that the model is correct, the observable mean reaction times follow an
approximate Normal distribution around the model prediction \(E(T_\tau)\) for
each condition. We can, therefore, use a standard goodness-of-fit measure
by \(z\)-standardization.

@@

In its current version \pkg{mathml} has some limitations. For example, it is
currently not possible to use the functions of \pkg{mathml} for writing
inline formulas, here, the user has to adopt the usual LateX notation.
Moreover, for long equations, we did not yet find a convenient way to insert
line breaks. This is mostly due to lacking support by MathML and LaTeX
renderers. For example, in its current stage, the LaTeX package `breqn` [@breqn]
is mostly a proof of concept.

@@

We assume that `nthroot(x, n)` is available in the current
namespace [manually defined, or from R package \CRANpkg{pracma}, @pracma], so that
the names of the arguments and their order are accessible to `canonical()` if
needed. As we can see below, \pkg{mathml} uses a default
representation `name(arguments)` for such unknown functions.

@@

A proper MathML term is obtained by `mlx/3` (the x in mlx indicates that it is
an extension and is prioritized over the default ml/3 rules). `mlx/3`
recursively invokes `ml/3` for translating the function arguments _X_ and _N_,
and then constructs the correct MathML entity `<mroot>...</mroot>`.

@@

# Acknowledgment

Supported by the Erasmus+ program of the European
Commission (2019-1-EE01-KA203-051708).
